Ultrafast Control of Spin-Spin Interactions in Two-Dimensional Magnetic Systems: Kane-Mele-Hubbard Model

### 2019-03-07 Paper/paper.tex
v5

\textit{Introduction}.--- Discovery of athermal all-optical control of order parameters in antiferromagnetic (AFM) and ferromagnetic (FM) materials by ultrashort, intense and high-frequency laser pulses has elevated spintronics to a new era of ultrafast magnetism \cite{kimel2005ultrafast,Rasing}. Although the underlying microscopic mechanism behind this phenomena is unclear yet, there have been many attempts to uncover its origins. Recently, it has been theoretically proposed that laser pulses can modify and even reverse the sign of exchange interaction \cite{PhysRevLett.113.057201,PhysRevLett.115.075301,Mentink2015,PhysRevLett.118.097701,Stepanov2017,Kitamura,MentinkReview,Barbeau2018}, the strongest spin interaction in magnetic systems. This finding has been supported by recent experiments \cite{Mikhaylovskiy2015,2018Natur.553..481G}. Another important spin interaction is Dzyaloshinskii-Moriya intercation (DMI), an antisymmetric exchange interaction appeared in inversion symmetry broken systems, that breaks the chiral symmetry of the system \cite{dzyaloshinsky1958, moriya1960,Alireza-DMI1, Alireza-DMI2}. While this interaction is much weaker than the exchange interaction, its presence has important consequences in magnetic materials such as weak ferromagnetism in AFM materials \cite{dzyaloshinsky1958, moriya1960}, stabilizing topological objects like chiral skyrmions \cite{2009Sci...323..915M,2017NatRM...217031F,AlirezaSk1, AlirezaSk2} and chiral domain walls \cite{Thiaville_2012,Parkin,AlirezaSW}, and exotic phase of topological magnon insulators \cite{Owerre2016, Elyasi2018, Chen2018, Owerre1, Owerre2}. The ratio between these two interactions control the tilting angle of canted spins. Proposing a mechanism to tune this ratio has important consequences in ultrafast spin dynamics and switching \cite{Mikhaylovskiy2015,Mentink2015,Stepanov2017,Kitamura}.

@@

To find renormalized spin Hamiltonian, we first drive an effective static Hamiltonian using the Floquet formalism \cite{Floquet1,Floquet2}. To do that, we transform the original time-dependent Hamiltonian, $\hat{H}(t)$, using a unitary transformation $\hat{U}(t) = e^{-i\hat{S}(t)}$,
\begin{equation}
\hat{H}'(t) = e^{i\hat{S}(t)} \left(\hat{H}(t)  -  i\partial_t \right) e^{-i\hat{S}(t)}.
\label{Htransformed}
\end{equation}
We perform the unitary transformation perturbatively in the hopping parameter. We can formally write $\hat{T}(t) = \eta \hat{T}(t)$, where $\eta$ play the role of a bookkeeping parameter in the perturbation expansion. We expand $\hat{S}(t) = \sum_\nu \eta^\nu \hat{S}^{(\nu)}(t)$ and $\hat{H}'(t) = \sum_\nu \eta^\nu \hat{H}'^{(\nu)}(t)$. Our demand is that the transformed Hamiltonian must have the same periodicity of the original time-dependent Hamiltonian thus the unitary transformation itself must have the same periodicity of the driven laser field so that $\hat{S}^{(\nu)}(t) = \sum_m e^{im\omega t}\hat{S}^{(\nu)}_m$. Additionally, we impose that the transformed Hamiltonian to be block diagonal in the doublon number $d$ while $\hat{S}(t)$ does not contain block-diagonal terms. With these conditions the unitary transformation might be uniquely determined,
\begin{equation}
\hat{S}^{(\nu)}(t) = \sum_{d \neq 0} \sum_m \eta^\nu \hat{S}^{(\nu)}_{d,m} e^{im\omega t},
\end{equation}
where $\hat{S}^{(\nu)}_{d,m}$ changes the double occupancy number by $d$. We expand the transformed Hamiltonian \ref{Htransformed} in power series of $\eta$ and determine $\hat{S}^{(\nu)}(t)$ iteratively in $\nu$ so that $\hat{H}'^{(\nu)}(t)$ is diagonal in the doublon number. After tedious but straightforward calculations, the transformed Hamiltonian up to the second order in the hopping parameter is obtained $\hat{H}'(t)= \hat{T}'(t)+\text{U}\hat{d}$, where
\begin{align}
\label{transformedH}
&\hat{T}'(t) \approx  - \sum_m \hat{T}_{0,m}(t)e^{im\omega t} + \n \\
&+ \frac{1}{2}\sum_{mn} \left( \frac{\left[\hat{T}_{1,n}, \hat{T}_{-1,m-n} \right]}{\text{U}+n\omega} - \frac{\left[\hat{T}_{-1,n}, \hat{T}_{1,m-n} \right]}{\text{U}-n\omega} \right) e^{im\omega t}.
\end{align}
Now, we are in the situation that we can calculate the effective static Hamiltonian by time averaging of the transformed Hamiltonian $\hat{H}_{\text{eff}}=\hat{P}_0\hat{H}'(t)\hat{P}_0$, where $\hat{P}_0$ is the projection operator to the $d=0$ subspace.
Doing some algebra, the effective static Hamiltonian in terms of creation and annihilation operators is obtained,
\begin{align}
\hat{H}_{\text{eff}} &= - \sum_{i,j, \tau, \tau'} \left(t_{ij}^{\tau} t_{ji}^{\tau'} \sum_{n} \frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega} \right)  \hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} \hat{c}_{j \tau'}^\dagger \hat{c}_{i \tau'}+\text{U}\hat{d}. \label{GeneralHeff}
\end{align}

Finally, the spin Hamiltonian, at half-filling and strong interaction limit of the effective Hamiltonian $\hat{H}_{\text{eff}}$, using the relations Eqs. (\ref{SpinOperatorInv1}) and (\ref{SpinOperatorInv2}), is given by,
\begin{align}
\label{MKMHeffw}
\tilde{H}_{\text{S}}(\omega) =& \tilde{J}_{1}\sum_{\langle i,j \rangle} \bs{S}_i\cdot\bs{S}_j + \tilde{J}_{2}\sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i\cdot\bs{S}_j, \n \\
&+ \sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i \tilde{\bs{\Gamma}}_{ij} \bs{S}_j +\sum_{\langle \langle i,j \rangle \rangle} \tilde{\bs{D}}_{ij}\cdot \bs{S}_i \times \bs{S}_j,
\end{align}
with the following renormalized spin interactions,
\begin{subequations}
\label{Renor-para}
\begin{align}
&\tilde{J}_{1(2),ij} = 2t_{1(2)}^2\frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega}, \\ 
&\tilde{\bs{\Gamma}}_{ij} = 2\Delta^2 \text{diag}(-1,-1,1) \frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega},\\ 
&\tilde{\bs{D}}_{ij} = - 4 t_2 \Delta \frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega} \nu_{ij} \hat{\mathrm{e}}_z. 
\end{align}
\end{subequations}
All spin interaction parameters are renormalized by the same function but $\alpha_{ij}$ is different for NN and NNN parameters. Thus, the ratios between these renormalized parameters are different than non-perturbed parameters which is a peculiar of honeycomb lattices described by the Kane-Mele-Hubbard model. The renormalized spin interactions presented in Eq. (\ref{Renor-para}) are helicity independent.

>>>>

\textit{Introduction}.--- Discovery of athermal all-optical control of order parameters in antiferromagnetic (AFM) and ferromagnetic (FM) materials by ultrashort, intense and high-frequency laser pulses has elevated spintronics to a new era of ultrafast magnetism \cite{kimel2005ultrafast,Rasing}. Although the underlying microscopic mechanism behind this phenomena is unclear yet, there have been many attempts to uncover its origins. Recently, it has been theoretically proposed that laser pulses can modify and even reverse the sign of exchange interaction \cite{PhysRevLett.113.057201,PhysRevLett.115.075301,Mentink2015,PhysRevLett.118.097701,Stepanov2017,Kitamura,MentinkReview,Barbeau2018}, the strongest spin interaction in magnetic systems. This finding has been supported by recent experiments \cite{2018Natur.553..481G,Mikhaylovskiy2015,Perakis,PhysRevLett.93.197403}. Another important spin interaction is Dzyaloshinskii-Moriya intercation (DMI), an antisymmetric exchange interaction appeared in inversion symmetry broken systems, that breaks the chiral symmetry of the system \cite{dzyaloshinsky1958, moriya1960,Alireza-DMI1, Alireza-DMI2}. While this interaction is much weaker than the exchange interaction, its presence has important consequences in magnetic materials such as weak ferromagnetism in AFM materials \cite{dzyaloshinsky1958, moriya1960}, stabilizing topological objects like chiral skyrmions \cite{2009Sci...323..915M,2017NatRM...217031F,AlirezaSk1, AlirezaSk2} and chiral domain walls \cite{Thiaville_2012,Parkin,AlirezaSW}, and exotic phase of topological magnon insulators \cite{Owerre2016, Elyasi2018, Chen2018, Owerre1, Owerre2}. The ratio between these two interactions control the tilting angle of canted spins. Proposing a mechanism to tune this ratio has important consequences in ultrafast spin dynamics and switching \cite{Mikhaylovskiy2015,Mentink2015,Stepanov2017,Kitamura}.

@@

To find renormalized spin Hamiltonian, we first drive an effective static Hamiltonian using the Floquet formalism \cite{Floquet1,2018NJPh...20i3022R,Floquet2}. To do that, we transform the original time-dependent Hamiltonian, $\hat{H}(t)$, using a unitary transformation $\hat{U}(t) = e^{-i\hat{S}(t)}$,
\begin{equation}
\hat{H}'(t) = e^{i\hat{S}(t)} \left(\hat{H}(t)  -  i\partial_t \right) e^{-i\hat{S}(t)}.
\label{Htransformed}
\end{equation}
We perform the unitary transformation perturbatively in the hopping parameter. We can formally write $\hat{T}(t) = \eta \hat{T}(t)$, where $\eta$ play the role of a bookkeeping parameter in the perturbation expansion. We expand $\hat{S}(t) = \sum_\nu \eta^\nu \hat{S}^{(\nu)}(t)$ and $\hat{H}'(t) = \sum_\nu \eta^\nu \hat{H}'^{(\nu)}(t)$. Our demand is that the transformed Hamiltonian must have the same periodicity of the original time-dependent Hamiltonian thus the unitary transformation itself must have the same periodicity of the driven laser field so that $\hat{S}^{(\nu)}(t) = \sum_m e^{im\omega t}\hat{S}^{(\nu)}_m$. Additionally, we impose that the transformed Hamiltonian to be block diagonal in the doublon number $d$ while $\hat{S}(t)$ does not contain block-diagonal terms. With these conditions the unitary transformation might be uniquely determined,
\begin{equation}
\hat{S}^{(\nu)}(t) = \sum_{d \neq 0} \sum_m \eta^\nu \hat{S}^{(\nu)}_{d,m} e^{im\omega t},
\end{equation}
where $\hat{S}^{(\nu)}_{d,m}$ changes the double occupancy number by $d$. We expand the transformed Hamiltonian \ref{Htransformed} in power series of $\eta$ and determine $\hat{S}^{(\nu)}(t)$ iteratively in $\nu$ so that $\hat{H}'^{(\nu)}(t)$ is diagonal in the doublon number. After tedious but straightforward calculations, the transformed Hamiltonian up to the second order in the hopping parameter is obtained $\hat{H}'(t)= \hat{T}'(t)+\text{U}\hat{d}$, where
\begin{align}
\label{transformedH}
&\hat{T}'(t) \approx  - \sum_m \hat{T}_{0,m}(t)e^{im\omega t} + \n \\
&+ \frac{1}{2}\sum_{mn} \left( \frac{\left[\hat{T}_{1,n}, \hat{T}_{-1,m-n} \right]}{\text{U}+n\omega} - \frac{\left[\hat{T}_{-1,n}, \hat{T}_{1,m-n} \right]}{\text{U}-n\omega} \right) e^{im\omega t}.
\end{align}
Now, we are in the situation that we can calculate the effective static Hamiltonian by time averaging of the transformed Hamiltonian $\hat{H}_{\text{eff}}=\hat{P}_0\hat{H}'(t)\hat{P}_0$, where $\hat{P}_0$ is the projection operator to the $d=0$ subspace.
Doing some algebra, the effective static Hamiltonian in terms of creation and annihilation operators is obtained,
\begin{align}
\hat{H}_{\text{eff}} &= - \sum_{i,j, \tau, \tau'} \left(t_{ij}^{\tau} t_{ji}^{\tau'} \sum_{n} \frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega} \right)  \hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} \hat{c}_{j \tau'}^\dagger \hat{c}_{i \tau'}+\text{U}\hat{d}. \label{GeneralHeff}
\end{align}

Finally, the spin Hamiltonian, at half-filling and strong interaction limit of the effective Hamiltonian $\hat{H}_{\text{eff}}$, using the relations Eqs. (\ref{SpinOperatorInv1}) and (\ref{SpinOperatorInv2}), is given by,
\begin{align}
\label{MKMHeffw}
\tilde{H}_{\text{S}}(\omega) =& \tilde{J}_{1}\sum_{\langle i,j \rangle} \bs{S}_i\cdot\bs{S}_j + \tilde{J}_{2}\sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i\cdot\bs{S}_j, \n \\
&+ \sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i \tilde{\bs{\Gamma}}_{ij} \bs{S}_j +\sum_{\langle \langle i,j \rangle \rangle} \tilde{\bs{D}}_{ij}\cdot \bs{S}_i \times \bs{S}_j,
\end{align}
with the following renormalized spin interactions,
\begin{subequations}
\label{Renor-para}
\begin{align}
&\tilde{J}_{1(2),ij} = 2t_{1(2)}^2\sum_n\frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega}, \\
&\tilde{\bs{\Gamma}}_{ij} = 2\Delta^2 \text{diag}(-1,-1,1) \sum_n\frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega},\\
&\tilde{\bs{D}}_{ij} = - 4 t_2 \Delta \sum_n\frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega} \nu_{ij} \hat{\mathrm{e}}_z.
\end{align}
\end{subequations}
All spin interaction parameters are renormalized by the same function but $\alpha_{ij}$ is different for NN and NNN parameters. Thus, the ratios between these renormalized parameters are different than non-perturbed parameters which is a peculiar of honeycomb lattices described by the Kane-Mele-Hubbard model. The renormalized spin interactions presented in Eq. (\ref{Renor-para}) are helicity independent.
